Problem of Step 4 (debias) of fGFPCA on NHANES data

### 2024-02-27 Docs/NhanesApplExample.Rmd
more exploration on reversed eigenvalue problem

set.seed(130)

library(here)
library(tidyverse)
library(ggpubr)
library(refund)
library(lme4)
library(ROCR)
library(GLMMadaptive)
library(kableExtra)
library(mvtnorm)
library(mgcv)
library(splines)
library(LaplacesDemon)
library(arsenal)
theme_set(theme_minimal())
```

@@

```{r}
mu <- predict(debias_glmm, type = "terms")[1:J, 1]
plot(1:J, mu, type = "l")
```

@@

```{r gen_data}
# generate score
xi <- rmvnorm(Ntr, mean = rep(0, K), sigma = diag(evals))
xi %>% data.frame() %>% pivot_longer(1:4, names_to = "PC", values_to = "score") %>%
  ggplot()+
  geom_boxplot(aes(x=PC, y=score))+
  geom_jitter(aes(x=PC, y=score), size = 0.5)

@@

Looks like FPCA did not capture the shape of PC2 and PC3.

Next I'd like to re-evaluate the eigenvalues

@@

Eigenvalues before and after re-evaluation

@@

At this point, I think we could say the reversed eigenvalue problem is an artifact of the dataset it self. For some reason, FPCA was not able to estimated PC2 and PC3 very well, and the GLMM model later on impose more penalty on the second random term.

>>>>

set.seed(227)

library(here)
library(tidyverse)
library(ggpubr)
library(refund)
library(lme4)
library(ROCR)
library(GLMMadaptive)
library(kableExtra)
library(mvtnorm)
library(mgcv)
library(splines)
library(LaplacesDemon)
library(arsenal)
library(RColorBrewer)
theme_set(theme_minimal())
```

@@

```{r}
mu <- predict(debias_glmm, type = "terms")[1:J, 1]
# plot(1:J, mu, type = "l", main = "Population mean")
plot(1:J, plogis(mu), type = "l", main = "Population mean on the probablity scale")
```

- Deviation from each PC from the population mean

Below are figures for $\hat{f}_0(t)+2\sqrt{\hat{\lambda}}_k\phi_l(k)$, but on the probablity scale

```{r}
df_pc_dev <- df_phi %>%
  mutate(mu = mu) %>%
  mutate(phi1_dev = mu + 2*sqrt(evals[1])*phi1,
         phi2_dev = mu + 2*sqrt(evals[2])*phi2,
         phi3_dev = mu + 2*sqrt(evals[3])*phi3,
         phi4_dev = mu + 2*sqrt(evals[4])*phi4)

df_pc_dev %>%
  pivot_longer(7:10) %>%
  mutate(value = plogis(value), mu = plogis(mu)) %>%
  ggplot()+
  geom_line(aes(x=sind, y=value))+
  geom_line(aes(x=sind, y=mu), col = "red")+
  facet_wrap(~name) 
```

It looks like subjects with high loadings on PC2 would be more active than population average during afternoon/night, and less active in the morning. PC3, on the other hand, seems to indicate more active at night and early morning and less active in midday. They both reflect more active at night time.

- Individual score

```{r}
data.frame(matrix(debias_glmm$coefficients[-c(1:9)], ncol = 4)) %>%
  pivot_longer(1:4) %>%
  ggplot(aes(x=name, y=value))+
  geom_boxplot()+
  geom_jitter(size = 0.2, alpha = 0.2)
```

Compared to the other three PCs, the distribution of $\hat{\xi}_{i3}$ seems to be shifted lower (positively skewed?).

@@

```{r gen_data}
# generate score
xi <- rmvnorm(Ntr, mean = rep(0, K), sigma = diag(evals))
# xi %>% data.frame() %>% pivot_longer(1:4, names_to = "PC", values_to = "score") %>%
#   ggplot()+
#   geom_boxplot(aes(x=PC, y=score))+
#   geom_jitter(aes(x=PC, y=score), size = 0.5)

@@

Next I'd like to re-evaluate the eigenvalues.

PS from the mgcv [user manual](https://cran.r-project.org/web/packages/mgcv/mgcv.pdf): *Note that discrete=TRUEmay result in re-ordering of variables in tensor product smooths for improved efficiency, and sp must be
supplied in re-ordered order*. Does this also apply to random effects? Does it have anything to do with the reversed eigenvalue functions? (I don't believe this is the case though, based on the experiments. Basciall I fit the model on a subset of 100 subjects setting discrete = F, and the smoothing parameters still flipped).

@@

- Eigenvalues before and after re-evaluation

@@

```{r, eval=FALSE}
new_df <- gen_df %>% filter(id == 1)
est_phi <- predict(debias_glmm2, newdata = new_df, type = "terms")[, 2:5] %>% 
  data.frame()
est_phi %>%
  mutate(sind=1:J) %>%
  pivot_longer(1:4) %>%
  ggplot()+
  geom_line(aes(x=sind, y= value))+
  facet_wrap(~name)
```

```{r eval=FALSE}
coef1 <- coef(debias_glmm2)[c("s(id):phi1.1", "s(id):phi2.1", "s(id):phi3.1", "s(id):phi4.1")]
coef1 <- matrix(coef1, nrow = J, ncol = 4, byrow = T)
(as.matrix(df_phi[, 2:5]) * coef1) %>% data.frame() %>%
  mutate(sind=1:J) %>%
  pivot_longer(1:4) %>%
  ggplot()+
  geom_line(aes(x=sind, y= value))+
  facet_wrap(~name)

```

At this point, I think we could say the reversed eigenvalue problem is an artifact of the dataset it self. It looks like FPCA captures the shape of eigenvalues very well, but for some reason, mgcv::bam thinks it is necessary to penalize the second eigenfunctions more than the third function. Why is that? Could it be that PC2 is more wiggly?

If we compare the eigenvalues used for data generation to the ones estimated by bam, we can see bam actually does a good job. But the order of eigenfunctions flipped, and that is because FPCA re-ordered the eigenfunctions based on eigenvalues.

```{r, class.source='fold-show'}
# generation
evals
# estimation
1/debias_glmm2$sp[2:5]
```
